Speeding up decimal multiplication

### 2020-11-25 paper/paper.tex
Update

\begin{procedure}[H]
\DontPrintSemicolon
\caption{MontgomeryReduce($a$)}
\KwData{
    Integer $\mu.$\;
    Integer $p$ coprime with $\mu.$\;
    Integer $p' \in [0; \mu-1]$ such that $p \cdot p' \equiv -1 \, (\mymod \, \mu).$
}
\KwIn{
    Integer $a \in [0; \mu p - 1].$
}
\KwOut{
    Integer $b \in [0; p-1]$ such that $b \equiv a \mu^{-1} \, (\mymod \, p).$
}
\Begin{
    $m \longleftarrow ((a \mymod \mu) \cdot p') \mymod \mu$\;
    $r \longleftarrow (a + m \cdot p) \mydiv \mu$
        \tcc*[r]{this division is exact} % apparently no need to use \; after \tcc
    \eIf{$r \ge p$}{
        \Return{$r - p$}\;
    }{
        \Return{$r$}\;
    }
}
\end{procedure}

@@

\begin{procedure}[H]
\caption{FindPrimesForN($n$, $p_{\max}, \ell$)}
\SetKw{KwAnd}{and}
\DontPrintSemicolon
\KwIn{$n, p_{\max}, \ell.$}
\KwOut{Set of $\ell$ largest, or less if the total number is less than $\ell,$ primes of form $p = c \cdot 3 \cdot 2^n + 1,$ where $p \le p_{\max}$ and $c$ is odd.}
\Begin{
    $r \longleftarrow \varnothing$\;
    $\psi \longleftarrow (p_{\max} - 1) \mydiv 2^n $\;
    \If{$\psi \mymod 2 = 0$}{
        $\psi \longleftarrow \psi - 1$\;
    }
    \While{$\psi \mymod 3 \ne 0$}{
        $\psi \longleftarrow \psi - 2$\;
    }
    \While{$\psi > 0$ \KwAnd $|r| < \ell$}{
        $p \longleftarrow \psi \cdot 2^n + 1$\;
        \If{$p$ is prime}{
            $r \longleftarrow r \cup \{p\}$\;
        }
        $\psi \longleftarrow \psi - 6$\;
    }
    \Return{$r$}\;
}
\end{procedure}

\begin{procedure}[H]
\caption{FindPrimes($w$, $n_{\min}$, $\ell$)}
\SetKw{Error}{error}
\DontPrintSemicolon
\KwIn{$w, n_{\min}, \ell.$}
\KwOut{Set of $\ell$ largest primes of form $p = c \cdot 3 \cdot 2^n + 1,$ where $p < 2^{w-1},$ $n \ge n_{\min},$ and $c$ is odd.}
\Begin{
    $r \longleftarrow \varnothing$\;
    $p_{\max} \longleftarrow 2^{w-1}-1$\;
    \For{$n \longleftarrow n_{\min}$ \KwTo $w-2$}{
        $r \longleftarrow r \cup \mathrm{FindPrimesForN}(n, p_{\max}, \ell)$\;
    }
    \eIf{$|r| < \ell$}{
        \Error{cannot find $\ell$ primes with required properties}\;
    }{
        \Return{$\ell$ largest values of $r$}\;
    }
}
\end{procedure}

@@

\begin{procedure}[H]
\caption{BitRevPermute($A$)}
\DontPrintSemicolon
\KwIn{Array $A$, $|A| = 2^n.$}
\KwOut{The result of applying the bit-reversal permutation to $A.$}
\Begin{
    $n \longleftarrow \log_2 |A|$\;
    \For{$k \longleftarrow 0$ \KwTo $2^n - 1$}{
        $\langle b_0, \ldots, b_{n-1} \rangle \longleftarrow$ bits of $k;$ $b_i \in \{0,1\}$ and $\sum_{i=0}^{n-1} b_i 2^i = k$\;
        $k' \longleftarrow \sum_{i=0}^{n-1} b_{n-i-1} 2^i$\;
        \If{$k < k'$}{
            {exchange $A[k]$ with $A[k']$}\;
        }
    }
    \Return{$A$}\;
}
\end{procedure}

@@

\begin{procedure}[H]
\caption{DecimationInTime($A$, $\omega$)}
\SetKw{KwBy}{by}
\DontPrintSemicolon
\KwIn{
    Array $A$ of field elements, $|A| = 2^n$.
    Field element $\omega = \xi^{\pm 1},$ where $\xi$ is a primitive $2^n$-th root of unity.
}
\KwOut{
    Array $A'$ such that $|A'| = |A|, A'[k] = \sum_{i=0}^{|A| - 1} A[i] \omega^{ik}.$
}
\Begin{
    $A \longleftarrow \mathrm{BitRevPermute}(A)$\;
    \For{$s \longleftarrow 1$ \KwTo $\log_2 |A|$}{
        $m \longleftarrow 2^s$\;
        $\omega_m \longleftarrow \omega^{|A| / m}$\;
        \For{$k = 0$ \KwTo $|A|-1$ \KwBy $m$}{
            $\varphi \longleftarrow \widehat{1}$\;
            \For{$j = 0$ \KwTo $m/2-1$}{
                $u \longleftarrow A[k + j]$\;
                $v \longleftarrow \varphi \cdot A[k + j + m/2]$\;
                $A[k + j] \longleftarrow u + v$\;
                $A[k + j + m/2] \longleftarrow u - v$\;
                $\varphi \longleftarrow \varphi \cdot \omega_m$\;
            }
        }
    }
    \Return{$A$}\;
}
\end{procedure}

@@

\begin{procedure}[H]
\caption{DecimationInFrequency($A$, $\omega$)}
\SetKw{KwBy}{by}
\SetKw{KwDownTo}{downto}
\DontPrintSemicolon
\KwIn{
    Array $A$ of field elements, $|A| = 2^n$.
    Field element $\omega = \xi^{\pm 1},$ where $\xi$ is a primitive $2^n$-th root of unity.
}
\KwOut{
    Array $A'$ such that $|A'| = |A|, A'[k] = \sum_{i=0}^{|A| - 1} A[i] \omega^{ik}.$
}
\Begin{
    \For{$s \longleftarrow \log_2 |A|$ \KwDownTo $1$}{
        $m \longleftarrow 2^s$\;
        $\omega_m \longleftarrow \omega^{|A| / m}$\;
        \For{$k = 0$ \KwTo $|A|-1$ \KwBy $m$}{
            $\varphi \longleftarrow \widehat{1}$\;
            \For{$j = 0$ \KwTo $m/2-1$}{
                $u \longleftarrow A[k + j]$\;
                $v \longleftarrow A[k + j + m/2]$\;
                $A[k + j] \longleftarrow u + v$\;
                $A[k + j + m/2] \longleftarrow \varphi \cdot (u - v)$\;
                $\varphi \longleftarrow \varphi \cdot \omega_m$\;
            }
        }
    }
    $A \longleftarrow \mathrm{BitRevPermute}(A)$\;
    \Return{$A$}\;
}
\end{procedure}

@@

The algorithm for in-place transposition of a square matrix is simple: element
$\langle i, j \rangle$ is exchanged with element $\langle j, i \rangle.$
The same approach works for transposing a square submatrix of a larger, generally rectangular
matrix.
For completeness, we provide the code below.

\begin{procedure}[H]
\caption{TransposeSquareSubMatrix($p$, $n$, $n'$)}
\DontPrintSemicolon
\KwData{
    Pointer $p$ to the first row, first column of $n$-by-$n$ submatrix of a possibly larger matrix
    with $n''$ columns and $n'$ rows stored in row-major order.
    (Formally, if the submatrix starts at $i$-th row, $0 \le i < n'$ and
    $j$-th column, $0 \le j < n'',$ then $p = q + n' \cdot i + j,$ where $q$ is the pointer to the
    beginning of the larger matrix.)

@@

For a cache-oblivious version of this algorithm, see~\cite{cache_square_mat_trans}.

@@

We provide the pseudocode for case of $C=2$ below.

Note it uses the routine $\mathrm{TransposeSquareSubMatrix}$.
It does not have to be implemented exactly in a way we demonstrated above;
it just needs to have the same semantics.
It can be implemented in a cache-friendly way by dividing the square submatrix into
tiles, perhaps recursively, transposing the elements inside tiles, and then swapping the tiles
themselves; see~\cite{cache_square_mat_trans} for details.

\begin{procedure}[H]
\caption{PermuteRho($p$, $\zeta$, $n$)}
\DontPrintSemicolon
\KwData{
    Pointer $p$ to array of size $2n$.
    Pointer $\zeta$ to scratch buffer of size $2n.$
    Integer $n.$
}
\KwResult{
    The permutation $\rho$ is applied to the array.
    The contents of the scratch buffer is undefined.
}
\Begin{
    \For{$i = 0$ \KwTo $2n-1$}{
        $\zeta[i] \longleftarrow p[i]$\;
    }
    \For{$i = 0$ \KwTo $n-1$}{
        $p[i] \longleftarrow \zeta[2i]$\;
        $p[i+n] \longleftarrow \zeta[2i+1]$\;
    }
}
\end{procedure}

\begin{procedure}[H]
\caption{UnPermuteRho($p$, $\zeta$, $n$)}
\DontPrintSemicolon
\KwData{
    Pointer $p$ to array of size $2n$.
    Pointer $\zeta$ to scratch buffer of size $2n.$
    Integer $n.$
}
\KwResult{
    The permutation $\rho^{-1}$ is applied to the array.
    The contents of the scratch buffer is undefined.
}
\Begin{
    \For{$i = 0$ \KwTo $2n-1$}{
        $\zeta[i] \longleftarrow p[i]$\;
    }
    \For{$i = 0$ \KwTo $n-1$}{
        $p[2i] \longleftarrow \zeta[i]$\;
        $p[2i+1] \longleftarrow \zeta[i+n]$\;
    }
}
\end{procedure}

\begin{procedure}[H]
\caption{TransposeMatrixNx2N($p$, $\zeta$, $n$)}
\DontPrintSemicolon
\KwData{
    Pointer $p$ to a matrix with $n$ rows and $2n$ columns stored in row-major order.
    Pointer $\zeta$ to scratch buffer of size $2n.$
    Integer $n.$
}
\KwResult{
    The matrix is transposed.
    The contents of the scratch buffer is undefined.
}
\Begin{
    \For{$i = 0$ \KwTo $n-1$}{
        $\mathrm{PermuteRho}(p + i \cdot 2n, \zeta, n)$\;
    }
    $\mathrm{TransposeSquareSubMatrix}(p, n, 2n)$\;
    $\mathrm{TransposeSquareSubMatrix}(p + n, n, 2n)$\;
}
\end{procedure}

\begin{procedure}[H]
\caption{TransposeMatrix2NxN($p$, $\zeta$, $n$)}
\DontPrintSemicolon
\KwData{
    Pointer $p$ to a matrix with $2n$ rows and $n$ columns stored in row-major order.
    Pointer $\zeta$ to scratch buffer of size $2n.$
    Integer $n.$
}
\KwResult{
    The matrix is transposed.
    The contents of the scratch buffer is undefined.
}
\Begin{
    $\mathrm{TransposeSquareSubMatrix}(p, n, 2n)$\;
    $\mathrm{TransposeSquareSubMatrix}(p + n, n, 2n)$\;
    \For{$i = 0$ \KwTo $n-1$}{
        $\mathrm{UnPermuteRho}(p + i \cdot 2n, \zeta, n)$\;
    }
}
\end{procedure}

>>>>

\begin{procedure}[!hbt]
\DontPrintSemicolon
\caption{MontgomeryReduce($a$)}
\KwData{
    Integer $\mu.$\;
    Integer $p$ coprime with $\mu.$\;
    Integer $p' \in [0; \mu-1]$ such that $p \cdot p' \equiv -1 \, (\mymod \, \mu).$
}
\KwIn{
    Integer $a \in [0; \mu p - 1].$
}
\KwOut{
    Integer $b \in [0; p-1]$ such that $b \equiv a \mu^{-1} \, (\mymod \, p).$
}
\Begin{
    $m \longleftarrow ((a \mymod \mu) \cdot p') \mymod \mu$\;
    $r \longleftarrow (a + m \cdot p) \mydiv \mu$
        \tcc*[r]{this division is exact} % apparently no need to use \; after \tcc
    \eIf{$r \ge p$}{
        \Return{$r - p$}\;
    }{
        \Return{$r$}\;
    }
}
\end{procedure}

@@

\begin{procedure}[!hbt]
\caption{FindPrimesForN($n$, $p_{\max}, \ell$)}
\SetKw{KwAnd}{and}
\DontPrintSemicolon
\KwIn{$n, p_{\max}, \ell.$}
\KwOut{Set of $\ell$ largest, or less if the total number is less than $\ell,$ primes of form $p = c \cdot 3 \cdot 2^n + 1,$ where $p \le p_{\max}$ and $c$ is odd.}
\Begin{
    $r \longleftarrow \varnothing$\;
    $\psi \longleftarrow (p_{\max} - 1) \mydiv 2^n $\;
    \If{$\psi \mymod 2 = 0$}{
        $\psi \longleftarrow \psi - 1$\;
    }
    \While{$\psi \mymod 3 \ne 0$}{
        $\psi \longleftarrow \psi - 2$\;
    }
    \While{$\psi > 0$ \KwAnd $|r| < \ell$}{
        $p \longleftarrow \psi \cdot 2^n + 1$\;
        \If{$p$ is prime}{
            $r \longleftarrow r \cup \{p\}$\;
        }
        $\psi \longleftarrow \psi - 6$\;
    }
    \Return{$r$}\;
}
\end{procedure}

\begin{procedure}[!hbt]
\caption{FindPrimes($w$, $n_{\min}$, $\ell$)}
\SetKw{Error}{error}
\DontPrintSemicolon
\KwIn{$w, n_{\min}, \ell.$}
\KwOut{Set of $\ell$ largest primes of form $p = c \cdot 3 \cdot 2^n + 1,$ where $p < 2^{w-1},$ $n \ge n_{\min},$ and $c$ is odd.}
\Begin{
    $r \longleftarrow \varnothing$\;
    $p_{\max} \longleftarrow 2^{w-1}-1$\;
    \For{$n \longleftarrow n_{\min}$ \KwTo $w-2$}{
        $r \longleftarrow r \cup \mathrm{FindPrimesForN}(n, p_{\max}, \ell)$\;
    }
    \eIf{$|r| < \ell$}{
        \Error{cannot find $\ell$ primes with required properties}\;
    }{
        \Return{$\ell$ largest values of $r$}\;
    }
}
\end{procedure}

@@

\begin{procedure}[!hbt]
\caption{BitRevPermute($A$)}
\DontPrintSemicolon
\KwIn{Array $A$, $|A| = 2^n.$}
\KwOut{The result of applying the bit-reversal permutation to $A.$}
\Begin{
    $n \longleftarrow \log_2 |A|$\;
    \For{$k \longleftarrow 0$ \KwTo $2^n - 1$}{
        $\langle b_0, \ldots, b_{n-1} \rangle \longleftarrow$ bits of $k;$ $b_i \in \{0,1\}$ and $\sum_{i=0}^{n-1} b_i 2^i = k$\;
        $k' \longleftarrow \sum_{i=0}^{n-1} b_{n-i-1} 2^i$\;
        \If{$k < k'$}{
            {exchange $A[k]$ with $A[k']$}\;
        }
    }
    \Return{$A$}\;
}
\end{procedure}

@@

\begin{procedure}[!hbt]
\caption{DecimationInTime($A$, $\omega$)}
\SetKw{KwBy}{by}
\DontPrintSemicolon
\KwIn{
    Array $A$ of field elements, $|A| = 2^n$.
    Field element $\omega = \xi^{\pm 1},$ where $\xi$ is a primitive $2^n$-th root of unity.
}
\KwOut{
    Array $A'$ such that $|A'| = |A|, A'[k] = \sum_{i=0}^{|A| - 1} A[i] \omega^{ik}.$
}
\Begin{
    $A \longleftarrow \mathrm{BitRevPermute}(A)$\;
    \For{$s \longleftarrow 1$ \KwTo $\log_2 |A|$}{
        $m \longleftarrow 2^s$\;
        $\omega_m \longleftarrow \omega^{|A| / m}$\;
        \For{$k = 0$ \KwTo $|A|-1$ \KwBy $m$}{
            $\varphi \longleftarrow \widehat{1}$\;
            \For{$j = 0$ \KwTo $m/2-1$}{
                $u \longleftarrow A[k + j]$\;
                $v \longleftarrow \varphi \cdot A[k + j + m/2]$\;
                $A[k + j] \longleftarrow u + v$\;
                $A[k + j + m/2] \longleftarrow u - v$\;
                $\varphi \longleftarrow \varphi \cdot \omega_m$\;
            }
        }
    }
    \Return{$A$}\;
}
\end{procedure}

@@

\begin{procedure}[!hbt]
\caption{DecimationInFrequency($A$, $\omega$)}
\SetKw{KwBy}{by}
\SetKw{KwDownTo}{downto}
\DontPrintSemicolon
\KwIn{
    Array $A$ of field elements, $|A| = 2^n$.
    Field element $\omega = \xi^{\pm 1},$ where $\xi$ is a primitive $2^n$-th root of unity.
}
\KwOut{
    Array $A'$ such that $|A'| = |A|, A'[k] = \sum_{i=0}^{|A| - 1} A[i] \omega^{ik}.$
}
\Begin{
    \For{$s \longleftarrow \log_2 |A|$ \KwDownTo $1$}{
        $m \longleftarrow 2^s$\;
        $\omega_m \longleftarrow \omega^{|A| / m}$\;
        \For{$k = 0$ \KwTo $|A|-1$ \KwBy $m$}{
            $\varphi \longleftarrow \widehat{1}$\;
            \For{$j = 0$ \KwTo $m/2-1$}{
                $u \longleftarrow A[k + j]$\;
                $v \longleftarrow A[k + j + m/2]$\;
                $A[k + j] \longleftarrow u + v$\;
                $A[k + j + m/2] \longleftarrow \varphi \cdot (u - v)$\;
                $\varphi \longleftarrow \varphi \cdot \omega_m$\;
            }
        }
    }
    $A \longleftarrow \mathrm{BitRevPermute}(A)$\;
    \Return{$A$}\;
}
\end{procedure}

@@

The algorithm for in-place transposition of a square matrix is simple: element
$\langle i, j \rangle$ is exchanged with element $\langle j, i \rangle.$
The same approach works for transposing a square submatrix of a larger, generally rectangular
matrix.
For completeness, we provide the code of $\mathrm{TransposeSquareSubMatrix}(p, n, n').$
For a cache-oblivious version of this algorithm, see~\cite{cache_square_mat_trans}.

\begin{procedure}[!hbt]
\caption{TransposeSquareSubMatrix($p$, $n$, $n'$)}
\DontPrintSemicolon
\KwData{
    Pointer $p$ to the first row, first column of $n$-by-$n$ submatrix of a possibly larger matrix
    with $n''$ columns and $n'$ rows stored in row-major order.
    (Formally, if the submatrix starts at $i$-th row, $0 \le i < n'$ and
    $j$-th column, $0 \le j < n'',$ then $p = q + n' \cdot i + j,$ where $q$ is the pointer to the
    beginning of the larger matrix.)

@@

@@

We provide the pseudocode for case of $C=2.$
Note it uses the routine $\mathrm{TransposeSquareSubMatrix}$.
It does not have to be implemented exactly in a way we demonstrated above;
it just needs to have the same semantics.
It can be implemented in a cache-friendly way by dividing the square submatrix into
tiles, perhaps recursively, transposing the elements inside tiles, and then swapping the tiles
themselves; see~\cite{cache_square_mat_trans} for details.

\begin{procedure}[!hbtp]
\caption{PermuteRho($p$, $\zeta$, $n$)}
\DontPrintSemicolon
\KwData{
    Pointer $p$ to array of size $2n$.
    Pointer $\zeta$ to scratch buffer of size $2n.$
    Integer $n.$
}
\KwResult{
    The permutation $\rho$ is applied to the array.
    The contents of the scratch buffer is undefined.
}
\Begin{
    \For{$i = 0$ \KwTo $2n-1$}{
        $\zeta[i] \longleftarrow p[i]$\;
    }
    \For{$i = 0$ \KwTo $n-1$}{
        $p[i] \longleftarrow \zeta[2i]$\;
        $p[i+n] \longleftarrow \zeta[2i+1]$\;
    }
}
\end{procedure}

\begin{procedure}[!hbtp]
\caption{UnPermuteRho($p$, $\zeta$, $n$)}
\DontPrintSemicolon
\KwData{
    Pointer $p$ to array of size $2n$.
    Pointer $\zeta$ to scratch buffer of size $2n.$
    Integer $n.$
}
\KwResult{
    The permutation $\rho^{-1}$ is applied to the array.
    The contents of the scratch buffer is undefined.
}
\Begin{
    \For{$i = 0$ \KwTo $2n-1$}{
        $\zeta[i] \longleftarrow p[i]$\;
    }
    \For{$i = 0$ \KwTo $n-1$}{
        $p[2i] \longleftarrow \zeta[i]$\;
        $p[2i+1] \longleftarrow \zeta[i+n]$\;
    }
}
\end{procedure}

\begin{procedure}[!hbtp]
\caption{TransposeMatrixNx2N($p$, $\zeta$, $n$)}
\DontPrintSemicolon
\KwData{
    Pointer $p$ to a matrix with $n$ rows and $2n$ columns stored in row-major order.
    Pointer $\zeta$ to scratch buffer of size $2n.$
    Integer $n.$
}
\KwResult{
    The matrix is transposed.
    The contents of the scratch buffer is undefined.
}
\Begin{
    \For{$i = 0$ \KwTo $n-1$}{
        $\mathrm{PermuteRho}(p + i \cdot 2n, \zeta, n)$\;
    }
    $\mathrm{TransposeSquareSubMatrix}(p, n, 2n)$\;
    $\mathrm{TransposeSquareSubMatrix}(p + n, n, 2n)$\;
}
\end{procedure}

\begin{procedure}[!hbtp]
\caption{TransposeMatrix2NxN($p$, $\zeta$, $n$)}
\DontPrintSemicolon
\KwData{
    Pointer $p$ to a matrix with $2n$ rows and $n$ columns stored in row-major order.
    Pointer $\zeta$ to scratch buffer of size $2n.$
    Integer $n.$
}
\KwResult{
    The matrix is transposed.
    The contents of the scratch buffer is undefined.
}
\Begin{
    $\mathrm{TransposeSquareSubMatrix}(p, n, 2n)$\;
    $\mathrm{TransposeSquareSubMatrix}(p + n, n, 2n)$\;
    \For{$i = 0$ \KwTo $n-1$}{
        $\mathrm{UnPermuteRho}(p + i \cdot 2n, \zeta, n)$\;
    }
}
\end{procedure}
